Subsidiary Recursion in Coq

### 2022-02-06 paper/paper.tex
more small edits

We need a fixed-point here because \verb|Alg| occurs in the definition
of \verb|Algf|.  This is an essential circularity, because we are
trying to express that algebras take in \verb|fold| functions, which
themselves may accept algebras.  The variable \verb|Alg| occurs
negatively in \verb|FoldT Alg R| which occurs negatively in
\verb|AlgF Alg X|.  Hence it occurs positively but not strictly
positively.  So we can indeed take a fixed-point of \verb|AlgF| to
define the constant \verb|Alg|.

@@

\begin{figure}
\begin{minted}{coq}
  Definition WordsByAlg(p : A -> bool)
    : Alg (ListF A) (Const (list (list A))) :=
    rollAlg (fun R fo wordsBy xs => 
      match xs with
        Nil => [] 
      | Cons hd tl =>
        if p hd then
          wordsBy tl
        else
          let (w,z) := breakr fo p tl in
            (hd :: w) :: wordsBy z
      end).
Definition wordsByr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                   (p : A -> bool)(xs : R) : list (list A) :=
  fo (Const (list (list A))) (FunConst (list (list A))) (WordsByAlg p) xs.
\end{minted}
\caption{Functions  \texttt{wordsBy} and \texttt{wordsByr}, and the algebra they
  fold (\texttt{WordsBy.v})}
\label{fig:wordsby}
\end{figure}

@@

To write this combinator using our infrastructure for subsidiary
recursion, we need to supply the mapped function with the fold
function for \verb|mapThrough|'s recursion.  This is so that the
mapped function can initiate a subsidiary recursion, returning a value
in the abstract type \verb|R| of \verb|mapThrough|'s recursion.  
The type we will use for mapped functions is:
\begin{minted}{coq}
Definition mappedT(A B : Set) : Set :=
  forall(R : Set)(fo:FoldT (Alg (ListF A)) R), A -> R -> B * R.
\end{minted}
\noindent This type is more informative than the Haskell type,
since it shows that the second component of the returned value
must have type \verb|R|, and hence must be (hereditarily) a tail
of the input.

>>>>

We need a fixed-point here because \verb|Alg| occurs in the definition
of \verb|Algf|.  This is an essential circularity, because we are
trying to express that algebras take in \verb|fold| functions, which
themselves may accept algebras.  The variable \verb|Alg| occurs
negatively in \verb|FoldT Alg R| which occurs negatively in
\verb|AlgF Alg X|.  Hence it occurs positively in \verb|Algf|, though not strictly
positively.  So we can indeed take a fixed-point of \verb|AlgF| to
define the constant \verb|Alg|.

@@

\begin{figure}
\begin{minted}{coq}
  Definition WordsByAlg(p : A -> bool)
    : Alg (ListF A) (Const (list (list A))) :=
    rollAlg (fun R fo wordsBy xs => 
      match xs with
        Nil => [] 
      | Cons hd tl =>
        if p hd then
          wordsBy tl
        else
          let (w,z) := breakr fo p tl in
            (hd :: w) :: wordsBy z
      end).
Definition wordsByr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                   (p : A -> bool)(xs : R) : list (list A) :=
  fo (Const (list (list A))) (FunConst (list (list A))) (WordsByAlg p) xs.
\end{minted}
\caption{The algebra \texttt{WordsByAlg}, and functions \texttt{wordsBy} and \texttt{wordsByr} folding it (\texttt{WordsBy.v})}
\label{fig:wordsby}
\end{figure}

@@

To write this combinator using our infrastructure for subsidiary
recursion, we will use this type for mapped functions:
\begin{minted}{coq}
Definition mappedT(A B : Set) : Set :=
  forall(R : Set)(fo:FoldT (Alg (ListF A)) R), A -> R -> B * R.
\end{minted}
\noindent This \verb|mappedT| type is more informative than the
Haskell type, since it shows that the second component of the returned
value must have type \verb|R|, and hence must be (hereditarily) a tail
of the input.  We need to supply mapped functions with the fold
function to use, which will come from \verb|mapThrough|'s recursion.
Mapped functions need this to initiate subsidiary recursions,
returning a value in the abstract type \verb|R| of \verb|mapThrough|'s
recursion.
